%% Tensorgrad: PLDI draft.
%% Build: latexmk -pdf main.tex   (acmart is in texlive 2023)
%% Submission target: PLDI (sigplan); evaluation written to also serve an
%% MLSys/NeurIPS-systems audience.
\documentclass[sigplan,10pt,review]{acmart}
%% For submission, switch to: \documentclass[sigplan,10pt,review,anonymous]{acmart}

\usepackage{booktabs}
\usepackage{listings}
\usepackage{xcolor}

\lstset{
  language=Python,
  basicstyle=\small\ttfamily,
  keywordstyle=\bfseries,
  commentstyle=\color{gray},
  columns=fullflexible,
  keepspaces=true,
}

%% Prominent TODO boxes so the draft is honest about what is landed vs pending.
\newcommand{\todopaper}[1]{\par\noindent\colorbox{orange!25}{%
  \parbox{\dimexpr\linewidth-2\fboxsep}{\textbf{TODO:} #1}}\par}

\setcopyright{none}
\settopmatter{printfolios=true}

\begin{document}

\title{Do the Calculus Once: A Tensor Language Where Equality Is
  Isomorphism and Differentiation Is Rewriting}

\author{Thomas D. Ahle}
\affiliation{\institution{}\country{}}
%% \author{...}  % TODO: author list

\begin{abstract}
Deep-learning frameworks differentiate programs by recording and replaying
tapes at runtime; computer-algebra systems differentiate expressions
symbolically but do not produce competitive kernels. We present
\textsc{tensorgrad}, a tensor language that does the calculus once,
ahead of time. Expressions are tensor networks with \emph{named edges} and
no axis order; equality is graph isomorphism, so commutativity,
associativity, and index renaming are quotiented out by representation
rather than by rewrite rules. Derivatives are rewrites: only irreducible
scalar leaves carry differentiation rules, and every composite---softmax,
layer normalization, attention---receives gradients and Hessians derived by
the language itself. A compiler lowers the resulting algebra to
straight-line PyTorch kernels. Its central pass restores reverse-mode
ordering to symbolically derived gradients, which naturally arrive
forward-mode-shaped: on a 12-block GPT-2--dimension gradient graph, planned
intermediates fall from 11.3\,exabytes to 19.8\,gigabytes (largest node
23\,MB). Every optimization rewrite is verified by exact modular
Schwartz--Zippel evaluation, and a value-numbering pass merges
algebraically equal subgraphs that no structural canonicalization can
identify, guarded by a fresh-seed refusal gate that turns any unsound merge
into a no-op. Derived backward passes reach the performance of
hand-written fused kernels on CPU---layer-norm, GELU, and softmax
gradients all track torch's built-in kernels to within measurement noise
despite being derived from primitive definitions---the
cross-entropy--softmax Hessian simplifies to its target-free two-term
closed form automatically, and on structured second-order workloads the
symbolic head start is decisive: the compiled Hessian of $x^{\top}Ax$
runs $8\times$ faster than \texttt{torch.func} and $20\times$ faster than
JAX. A three-block GPT
trains to 100\% task accuracy with autograd disabled globally---forward,
all 53 gradients, and the AdamW optimizer compiled into a single fused
program. All measurements are ahead-of-time CPU compilation.
\end{abstract}

\maketitle

\section{Introduction}\label{sec:intro}

This work began with a complaint. A researcher wanted the Hessian of
softmax cross-entropy---not to train a network, but to \emph{understand}
one---and found that no mainstream framework would give it to them in a
form a human could read. Automatic differentiation frameworks
\cite{jax,pytorch} will evaluate Hessian--vector products efficiently, but
the Hessian itself, as an object with visible structure
($\mathrm{diag}(s) - ss^{\top}$, independent of the target when the target
lies on the simplex), is available only by hand derivation. Computer
algebra systems \cite{mathematica,sympy} can derive it, but they neither
speak tensors with named indices natively nor compile their results into
kernels one could train with.

\textsc{Tensorgrad} closes this gap with a two-layer design that we argue
is the right division of labor for symbolic tensor computation:

\begin{itemize}
\item \textbf{The language layer normalizes.} Expressions are tensor
  networks; simplification is deterministic, search-free, semantics-%
  preserving rewriting toward a normal form. There is no cost model and no
  backtracking at this layer.
\item \textbf{The compiler layer optimizes.} Everything that requires
  search, cost models, or randomized reasoning---contraction ordering,
  reverse-mode recovery, numerical stabilization, layout assignment,
  kernel selection---lives in compiler passes over a lowered IR, verified
  independently of the language's notion of equality.
\end{itemize}

Mainstream systems blur these layers: \texttt{sympy}'s
\texttt{simplify} is a heuristic search with a complexity objective;
Mathematica's \texttt{FullSimplify} explores rule sequences under a time
budget; XLA \cite{xla} and TorchInductor \cite{pytorch2} optimize numerics with no algebraic layer
above them. Keeping the layers separate is what lets the language stay
small---the entire user-facing kernel is nine node types and one equality
relation---while the compiler grows ambitious.

\paragraph{Contributions.}
\begin{enumerate}
\item A tensor language in which \emph{equality is graph isomorphism}
  (\S\ref{sec:language}). Axes are named edges; expressions never specify
  an axis order; commutativity, associativity, and $\alpha$-renaming are
  unexpressible rather than rewritten away. Equality is decided by
  compositional canonical fingerprints---exact on the tree-shaped parts of
  expressions, Weisfeiler--Leman--refined on contraction multigraphs, with
  exact verification reserved for the rare fingerprint collisions
  (\S\ref{sec:iso}). Value constraints are \emph{equations in the language
  itself} with a termination order (\S\ref{sec:constraints}).
\item A differentiation discipline in which only irreducible leaves carry
  derivative rules (\S\ref{sec:autodiff}). We show empirically that a
  user-defined softmax---three lines of primitives---receives a correct
  gradient, a fused and numerically stable compiled kernel, and a
  closed-form Hessian, with no framework code (\S\ref{sec:tier0}).
\item An \emph{adjoint pass} that restores reverse-mode ordering to
  symbolically derived gradients (\S\ref{sec:adjoint}). We identify why no
  local cost model can find reverse mode (a cost valley: dissolving one
  chain boundary improves the objective by ${<}2\%$ until the entire chain
  collapses), and give a scoring-free structural algorithm. Planned
  gradient intermediates on transformer graphs drop by nearly nine orders
  of magnitude (Table~\ref{tab:adjoint}).
\item Randomized algebraic verification as compiler infrastructure
  (\S\ref{sec:sz}): IR nodes evaluate exactly over $\mathbb{F}_p$ at
  random points (Schwartz--Zippel \cite{schwartz1980,zippel1979,demillo1978}), and a
  consolidation pass merges value-equal subgraphs modulo axis permutation
  under a \emph{fresh-seed refusal gate} that converts any unsound merge
  into a clean fallback. On a 3-block GPT training step this removes 71\%
  of compute nodes and $4.5\times$ of runtime.
\item Technology mapping \cite{dagon} for a language without axis order
  (\S\ref{sec:techmap}): because programs never over-specify layout, the
  compiler assigns physical axis orders globally, reducing every remaining
  transpose to a zero-cost view, and maps contractions onto BLAS cells.
\item An evaluation (\S\ref{sec:eval}) showing derived backward passes
  reaching hand-written fused-kernel performance on CPU, automatic
  recovery of hand-derived Hessian structure, and an end-to-end GPT trained with
  \texttt{torch.set\_grad\_enabled(False)}: forward, all gradients, and
  AdamW as one compiled program.
\end{enumerate}

The comparison with JAX runs through the paper and is honest in both
directions: \texttt{jax.grad} composes and compiles \emph{numerics},
and XLA's whole-program fusion wins the training-step wall clock on CPU
at every scale we measured (\S\ref{sec:e2e}); tensorgrad simplifies
\emph{algebra} before compiling, which is where target-free Hessians,
structural zeros, and the $8$--$20\times$ second-order wins of
\S\ref{sec:hessian} come from.

\section{The Language}\label{sec:language}

\subsection{Named edges and tensor networks}

A \textsc{tensorgrad} expression is a tensor network \cite{penrose1971,tensorcookbook}: a multigraph whose
vertices are tensors and whose edges are named indices. A variable
declares its edges and their (symbolic) sizes:

\begin{lstlisting}
b, s, v, d = symbols("batch seq vocab d")
X  = Variable("X", batch=b, seq=s, d=d)
W  = Variable("W", d=d, vocab=v)
logits = X @ W          # contracts the shared edge `d`
\end{lstlisting}

There is no axis order: \texttt{X} is not a ``$b \times s \times d$
array'' but a tensor with three named edges. Contraction is edge fusion;
\texttt{@} contracts edges with equal names. Multi-head attention needs no
\texttt{view}/\texttt{transpose}: heads are an edge named \texttt{head},
and attention keys are made by renaming \texttt{seq} to \texttt{key}.
Order-dependence is unexpressible, which has a consequence we exploit in
\S\ref{sec:techmap}: the compiler owns physical layout globally, because
the program never gave that freedom away.

The language kernel has nine node types: \texttt{Variable},
\texttt{Delta} (Kronecker/copy tensors, which also serve as hyperedges),
\texttt{Zero}, \texttt{Sum} (weighted, unordered), \texttt{Product}
(unordered), \texttt{Function} (opaque signatures over inputs),
\texttt{Derivative}, \texttt{Rename}, and constants. The type set is
closed by design; operations over it grow (\S\ref{sec:rules-as-data}).

\subsection{Equality is isomorphism}\label{sec:iso}

Two expressions are equal iff their tensor networks are isomorphic
(respecting labels; optionally matching free-edge names). This is a
semantic decision baked into the representation: since \texttt{Product}
holds an unordered multiset of factors wired by named edges,
commutativity and associativity cannot be stated, and $\alpha$-renaming of
contracted edges is invisible. Where \texttt{sympy} canonicalizes
\texttt{Mul} arguments by sorting and Mathematica matches modulo
\texttt{Orderless} attributes, \textsc{tensorgrad} has nothing to do: the
quotient is structural.

Deciding isomorphism is, of course, not free. We use a compositional
canonical fingerprint, computed bottom-up and cached on immutable nodes:
a node's fingerprint hashes its label and its children's fingerprints
(sorted where children are interchangeable, slotted where they are not),
plus a Weisfeiler--Leman \cite{weisfeiler1968} refinement of the contraction wiring for
\texttt{Product} multigraphs. Three properties matter:

\begin{itemize}
\item \emph{Soundness of rejection}: isomorphic terms have equal
  fingerprints, so unequal fingerprints prove inequality. Real workloads
  are rejection-dominated (hash-consing, memoization, term collection).
\item \emph{Exactness on trees}: outside \texttt{Product}, expressions are
  trees, where the fingerprint is the classic AHU canonical code \cite{aho1974}---not a
  heuristic.
\item \emph{Acceptance}: equal refined fingerprints are accepted directly
  (invariant~I1: an equal refined fingerprint exhibits an edge-preserving
  isomorphism, sound up to a 64-bit hash collision), with an exact
  isomorphism check (VF2 \cite{vf2}; cf.\ nauty \cite{nauty}) held in
  reserve for the coarse-collision case. Because product vertices arrive
  heavily colored (names, shapes, child fingerprints), refinement
  shatters essentially all practical instances: across a full 3-block GPT
  build (loss + 53 gradients, $8{,}736$ equality queries reaching the
  isomorphism routine), the exact VF2 fallback fired \textbf{zero times}.
  Graph-isomorphism hardness is not repealed; on real workloads it is
  never reached.
\end{itemize}

The fingerprint replaced a networkx-based \cite{networkx} path (rebuild the graph, run WL,
per query, per node) that spent its time constructing graphs; the
compositional design is what makes hashing $O(\text{degree})$ per new node
rather than $\Omega(\text{subtree})$, a distinction that decides whether
deep-model simplification is linear or quadratic
(\S\ref{sec:depth}).

\paragraph{One source of structural truth.}
Both the fingerprint and the verification graph must agree on what
participates in identity (symmetries, value constraints, function
parameters). We derive both as \emph{folds over a single declarative
description}: each node type declares
$\textit{structure}() = (\textit{label},\ \textit{children},\
\textit{child roles},\ \textit{junctions})$, where equal child roles
declare interchangeability and a \emph{junction} is a set of endpoints
identified as one wire. Contraction, broadcast pass-through, hyperedges,
and edge exposure are all junctions; inner edge names are used only to
group endpoints and then discarded, making $\alpha$-invariance a property
of the datum rather than a discipline of two encoders.

The refactor is complete: both the fingerprint and the verification
graph are now folds over \textit{structure}() declarations, and the two
legacy encoders were deleted after bridge-testing agreement over a
randomized corpus.

\subsection{Constraints are equations}\label{sec:constraints}

Facts about a variable's \emph{values} are declared as equations in the
language itself:

\begin{lstlisting}
y = Variable("y", batch=b, vocab=v)
y = y.with_eq_constraint(F.sum(y, ["vocab"]), 1)   # rows sum to 1
W = W.with_eq_constraint(W @ W.rename(i="j"),
                         Delta(i, "i", "j"))       # orthonormal
\end{lstlisting}

During simplification, any subnetwork isomorphic to a declared left-hand
side is replaced by the right-hand side. Because tensorgrad's own
canonicalization turns Hadamard squares into \texttt{pow} nodes and sums
into ones-contractions, simplex, unit-norm, and orthogonality constraints
are all two-factor patterns matched in the existing pairwise rewrite loop.
Termination is a declaration-time obligation: the right-hand side must be
strictly smaller (node count) than the left, so constraint rewriting
always shrinks. Equation sets must be closed under the variable's symmetry
orbits (checked at declaration). Constraints participate in identity but
deliberately \emph{not} in differentiation: gradients are taken in the
ambient space.

The payoff is visible in \S\ref{sec:hessian}: declaring that targets lie
on the simplex is exactly what makes the cross-entropy--softmax Hessian
simplify to a target-free closed form.

\subsection{Rules as code, becoming rules as data}\label{sec:rules-as-data}

The expression-problem split is in place: the data-type module holds
nine node types and their structural declarations, while simplification
lives in an enumerable rule catalog with a single engine; the equation
engine of \S\ref{sec:constraints} was its first rules-as-data component,
carrying its own termination order. The same movement recurs in the
backend, where kernel mappings are cells in a registry rather than
compiler code (\S\ref{sec:techmap}).

\todopaper{\textbf{Confluence harness}: a declared
termination measure per rule plus randomized local-confluence testing
(overlapping redexes, both application orders, canonical-equality of
normal forms, Schwartz--Zippel as the semantic backstop separating
confluence bugs from soundness bugs); by Newman's lemma \cite{newman1942} this tests exactly
the property required. Report the number of rule pairs fuzzed and any
divergences found.}

\section{Differentiation as Rewriting}\label{sec:autodiff}

\texttt{Derivative} is a node. Simplification pushes it through the
algebra: linearity over \texttt{Sum}, the product rule over
\texttt{Product} (new edges appear for the differentiated variable's
indices), and the chain rule at \texttt{Function} nodes, whose signatures
supply derivatives \emph{only at irreducible leaves}: \texttt{exp},
\texttt{log}, \texttt{pow}, \texttt{tanh}, \texttt{erf}, matrix inverse,
determinant, sign. Everything else is a composition and differentiates by
rewriting. This is a design law, not a default: the system contains no
hand-written Jacobian for any composite.\footnote{The law was violated
twice during development---hand-fused softmax and gather Jacobians added
for performance---and both violations were later deleted with zero
measured regression: the derived gradients compile to identical programs
(\S\ref{sec:eval}). We keep this footnote as an honest ablation: the
hand-derived math was redundant, but discovering that required the
compiler passes of \S\ref{sec:compiler} to exist. A separate, opt-in
tier of technology-mapping \emph{cells} (\S\ref{sec:techmap}) does map
selected composites---attention, layer norm, GELU---onto fused
forward+backward library kernels; those are not derivative rules in the
language: the derived gradient remains the specification, and each
cell's backward is machine-checked against it.}

Integer-indexed operations fit the same discipline: \texttt{gather} is
a contraction with a one-hot indicator of integer data
($\texttt{table} \mathbin{@} \texttt{one\_hot}(\texttt{idx})$), linear in
the table, so its scatter-add gradient falls out of the product rule.

\subsection{The depth problem}\label{sec:depth}

Symbolic systems traditionally fail on deep models for a reason
independent of algebra: expression DAGs with residual sharing
($x \mapsto x + f(x)$) have exponentially many root-to-leaf \emph{paths},
and any recursion that walks paths---substitution, dependency queries,
renaming, unmemoized simplification---is exponential in depth even though
the DAG is small. We treat this as a single disease with a single cure:
all structural recursions are memoized per $(\text{node},
\text{query})$ pair, and renaming is \emph{lazy}: \texttt{Rename} is a
node, constructed in $O(1)$, lowered by the same simplification pass that
owns all other rewriting. With these in place, building and normalizing
the loss and all 53 parameter gradients of a 3-block GPT takes under a
second (\S\ref{sec:eval} reports the one-time compile cost).

\subsection{Tier 0: what a user must write}\label{sec:tier0}

The founding promise is that a researcher writes mathematics and nothing
else. Concretely, with the library's softmax ignored:

\begin{lstlisting}
def my_softmax(x, dim):
    e = F.exp(x)
    return e / F.sum(e, dim, keepdims=True)

loss = F.sum(y * -F.log(my_softmax(x, ["v"])))
step = compile_to_callable(loss, loss.grad(x))
\end{lstlisting}

Measured, from these lines alone: the gradient matches
\texttt{torch.autograd} to $3\times10^{-8}$; the generated kernel
contains \texttt{torch.softmax} and \emph{no} \texttt{torch.exp}
(the stabilization pass of \S\ref{sec:stabilize} matches IR shapes, not
names); the compiled program is finite at $|{\log its}|\approx 600$; and
$\texttt{loss.grad(x).grad(x)}$---the Hessian that motivated the
project---derives and compiles in 0.05\,s.

\section{The Compiler}\label{sec:compiler}

The compiler is a fixed pipeline of IR-to-IR passes
(Figure~\ref{fig:pipeline}); all outputs share one hash-consed DAG, so
every pass sees the loss and all gradients at once.

\begin{figure}
\begin{verbatim}
symbolic tensors
  | normalize    resolve Derivative nodes (memoized,
  |              shared across all outputs)
  v
  lower         one hash-consed einsum DAG for ALL
  |             outputs; Deltas become shared wires;
  |             structured tensors -> affine rows
  v
  adjoint       restore reverse-mode ordering (S4.2)
  v
  factor        (un)distribute by contraction cost
  v
  stabilize     exp-ratio / logsumexp -> stable kernels
  v
  consolidate   SZ value numbering mod permutation (S5)
  v
  layout        global physical axis assignment (S6)
  v
  codegen       BLAS cells, index_select/index_add,
  |             spec-time strides, liveness `del`s
  v
  runtime       per-shape specialization; optional
                torch.compile (fullgraph=True)
\end{verbatim}
\caption{The pass pipeline. All outputs (loss and every gradient) share
one DAG, so forward work is shared across outputs by construction.}
\label{fig:pipeline}
\end{figure}

\subsection{Lowering: structure as equations, deltas as wires}

Products lower to einsum nodes over positional wires; named edges survive
only at program boundaries. \texttt{Delta} tensors never materialize:
they alias wires via union--find. Structured tensors (convolution,
reshape, shift, one-hot) are indicators of integer-affine index relations
$[\sum_i c_i e_i = k]$, carried as constraint rows on einsum nodes and
eliminated at code generation into strided views (\texttt{as\_strided},
\texttt{index\_select}); a dense indicator remains as the always-correct
fallback, so structure is a fast path, never a correctness requirement.

\subsection{The adjoint pass: recovering reverse mode}\label{sec:adjoint}

Symbolic differentiation produces \emph{forward-mode-shaped} gradients:
the derivative of a deep composition arrives as Jacobian chains that drag
the parameter's indices through every layer above it. The numbers are
disqualifying (Table~\ref{tab:adjoint}): one transformer block at GPT-2 \cite{gpt2}
dimensions plans 404{,}597\,GB of intermediates for three gradients; the
same mathematics, associated cotangent-first, needs megabytes.

The key diagnosis is that \emph{no local cost model can fix this}. The
factoring pass's greedy sweeps, which successfully recover shared
backward structure elsewhere, cannot cross the gap: dissolving a single
boundary of a Jacobian chain improves local cost by under $2\%$ (measured
$0.2\%$ at the chain head), because the neighboring node is still read at
full width---the payoff appears only when the entire chain has collapsed.
Greedy search faces a valley with the reward at the far side.

The adjoint pass is therefore \emph{scoring-free and structural}. A node
is a \emph{transient Jacobian} if its element count exceeds every program
input and output by a margin; such nodes must not survive. Two rules run
to a fixpoint: (R)~at a small-output einsum reading an inflated operand,
contract everything else down to the shared interface---an explicit,
hash-consed \emph{cotangent} node---then dissolve one boundary of the
inflated operand's chain; (M)~cotangents reading the same inflated node
accumulate: per-node adjoint accumulation across sums, across gradient
outputs, program-wide. Rule M's globality is what carries the pass across
arbitrary depth: contributions from different stacked blocks meet at the
shared node exactly as in classical reverse-mode AD \cite{griewank2008}, but discovered as an
IR transformation over symbolically derived algebra. Every rewrite is
einsum reassociation or distribution, pinned by the verification of
\S\ref{sec:sz}.

\begin{table}
\caption{Adjoint pass: planned gradient intermediates (fp32) before
lowering-only, and after the full pipeline. ``max'' is the largest single
node. CPU, symbolic shapes fixed at GPT-2 dimensions
($d{=}768$, $h{=}12$, $s{=}256$, $b{=}8$) except gpt-nano ($d{=}48$).}
\label{tab:adjoint}
\small
\begin{tabular}{lrrr}
\toprule
program & lowered & final & max node \\
\midrule
1 block, loss+3 grads & 404{,}597\,GB & 0.87\,GB & 25\,MB \\
3-block nano, loss+53 grads & 841\,GB & 0.41\,GB & 1.0\,MB \\
12 blocks, loss+3 grads & 11.26\,EB & 19.8\,GB & 23\,MB \\
\bottomrule
\end{tabular}
\end{table}

The 3-block program \emph{runs}: 185\,ms per step eager on two CPU
threads, every gradient within $7\times10^{-6}$ relative of
\texttt{torch.autograd}. One measurement deserves a remark: the
attention key-bias gradients agree with autograd only under an absolute
tolerance floor---because their true value is \emph{exactly zero} by
softmax shift-invariance. The symbolic system knows a theorem the numeric
comparison must be protected from.

\subsection{Stability is a compiler contract}\label{sec:stabilize}

Derived gradients arrive as expanded primitive algebra
($\exp/\Sigma\exp$ ratios). A stabilization pass over the IR recognizes
these shapes---including inside gradients and Hessians, and including
numerator/denominator pairs that factoring has separated---and emits
fused stable kernels (\texttt{softmax}, \texttt{log\_softmax},
\texttt{logsumexp}). The pass matches shapes, not names, which is what
makes \S\ref{sec:tier0} work for user-defined functions. The same
contract covers rational shapes: $\log(1+e^{x})$ maps to
\texttt{softplus}, and its \emph{derivative}---which arrives as
$e^{x}/(1+e^{x})$ with the $e^{x}$ smeared through the derivative's
diagonal einsum---is matched through that einsum and maps to
\texttt{sigmoid}, so softplus training is exact at $x{=}100$ where the
literal composition returns \texttt{inf} forward and \texttt{nan}
backward. Differentiation happened symbolically upstream, so the
rewrite only ever touches surviving forward compositions. Boundary,
measured: a shift-invariant softmax variant
($1/\sum_j e^{x_j - x_i}$) is \emph{not} matched---and compiles to its
literal (correct, and in this case self-stabilizing) algebra. Misses
degrade performance, never correctness. The eager interpreter runs no
IR passes and is deliberately simple; compiled programs are the stability
contract.

\section{Randomized Verification and Consolidation}\label{sec:sz}

Structural canonicalization cannot see \emph{algebraic} equality: after
factoring, the 53 gradients of a 3-block GPT contain hundreds of cotangent
chains that compute equal tensors through different contraction groupings
and axis orders. Measured: a full recursive quotient modulo axis
permutation removes only 568 of 2{,}925 compute nodes; the equality is
semantic, not syntactic.

We therefore verify and merge \emph{by value}. Every IR node can be
evaluated exactly over $\mathbb{F}_p$ (Mersenne $p$) at seeded random
points, with transcendental atoms mapped to fresh field elements keyed by
the value-fingerprint of their inputs---so $\tanh(x)^{\top}$ and
$\tanh(x^{\top})$ collide, as they must, while $\exp$ never accidentally
equals $\log$. This machinery serves twice:

\begin{itemize}
\item \emph{Rewrite verification}: rewrite classes are pinned by
  fingerprint equality before/after in the test suite (a
  Schwartz--Zippel bound \cite{schwartz1980,zippel1979} quantifies the
  error probability), alongside a differential property harness that
  compares the compiler against a deliberately simple interpreter on
  hundreds of random expressions.
\item \emph{Consolidation}: value numbering \cite{gulwani2004gvn} modulo axis permutation
  merges nodes whose evaluations agree at every trial under one
  consistent permutation; consumers are rewired through zero-cost
  permuted views. A \emph{fresh-seed refusal gate} then re-evaluates the
  original and consolidated programs side by side at new points and
  rejects the entire rewrite on any mismatch: a bookkeeping bug becomes a
  no-op, never a miscompile.
\end{itemize}

A postmortem argues this gate should guard \emph{every} pass, not just
consolidation. The longest-lived wrong-value bug in the compiler's
history was in factoring: splicing a child einsum whose broadcast-only
output wire the parent contracts left that wire referenced by nothing,
and the einsum constructor silently dropped it---losing an exact factor
of $\dim(w)$, i.e.\ compiling $\sum_a (x \otimes \mathbf{1}_a)$ to
$x$ instead of $|a|\,x$. The differential harness caught it; the
refusal gate would have refused it \emph{at compile time}, but
factoring ran outside the gate. The fix is the same accounting the
pass's own absorption step already did for its orphan wires---and the
lesson is that randomized verification is cheap enough to be the
default contract between passes, not a special feature of one.

On the 3-block GPT training step, consolidation removes
$2{,}925 \rightarrow 844$ compute nodes ($7{,}127 \rightarrow 2{,}052$
emitted kernels) and reduces the step from 185\,ms to 41\,ms eager. We
believe the broader point is of independent interest: randomized
polynomial identity testing is practical \emph{compiler} infrastructure,
not just a proof technique.

\section{Technology Mapping}\label{sec:techmap}

The analogy that organizes the backend is physical design in EDA: a
technology-independent netlist (the einsum DAG) is mapped onto a
characterized cell library (BLAS \cite{blas} and fused kernels) \cite{dagon}, and placement
(layout) is a global optimization variable.

\paragraph{Global layout assignment.} Because the language never fixes an
axis order, every IR node's physical layout is free. A reverse-topological
voting pass lets cells that care (matmuls, reductions) propose operand
layouts, elementwise nodes propagate them, and inputs pin to declared
order; BLAS absorbs one transpose per operand for free, so only genuine
copies count as cost. Result on the GPT program: \textbf{zero}
permute-copies---every remaining \texttt{permute} in generated code is a
stride view absorbed by its consumer. This is the concrete return on
under-specification: a PyTorch program has already committed to layouts
everywhere; a tensorgrad program has committed nowhere.

\paragraph{Cell selection.} Two-operand einsums that are (batched)
matmuls emit \texttt{mm}/\texttt{bmm}/\texttt{addmm} with merged reshape
groups (einsum count halved on the GPT program); one-hot contractions
emit \texttt{index\_select}/\texttt{index\_add\_}; reductions and
Hadamard-reductions emit fused forms. The generated code is
straight-line, with spec-time integer strides (it traces as a single
dynamo graph, so \texttt{torch.compile(fullgraph=True)} succeeds and
Inductor serves as a whole-program kernel generator), and liveness
\texttt{del}s that cut peak memory by up to $3\times$.
Not every EDA transplant survives measurement: a shared-operand GEMM
batching pass (the $q/k/v$ projections read the same activation; stack
the weights, run one GEMM, hand out views) is value-exact but
\emph{regresses the GPT step by $69\%$ on CPU}---\texttt{stack} is a
per-step copy because weights are step inputs, and a batched small GEMM
lowers to a slice loop rather than one BLAS call. It ships default-off.
We report this because the methodology is the point: every pass here is
gated on measured wall clock, not on launch-count intuition.

\paragraph{The cell library is data.} Every fused kernel is one object
in one registry (\texttt{compiler/cells.py}), and the registry spans four
different relationships to differentiation: \emph{differentiable leaves}
(matrix inverse and determinant, whose cookbook derivative rules expand
symbolically upstream, the cell mapping only surviving forward
applications onto \texttt{torch.linalg}); \emph{opaque forward+reverse
pairs} (attention, layer norm, GELU \cite{attention,layernorm,gelu}: a reverse-VJP hook emits a fused
backward cell, machine-verified against the derived gradient);
a \emph{multi-output, non-differentiable} optimizer update (AdamW \cite{adamw}:
$w',m',v'$ from one call, no VJP -- making it opaque also removed a
duplicated backward that value numbering could not see through, a
$1.5\times$ step improvement); and \emph{rewrite patterns carried by the
cells themselves}, applied by one generic engine:
$\mathrm{inverse}(A)\,b$ becomes \texttt{linalg.solve}, and
$\log\det A$ becomes \texttt{slogdet} -- the Gaussian-process example
runs in plain float32 because the latter rescues a determinant that
underflows to exactly $0$. The cells also \emph{share work}: when one
operand feeds two or more of $\{\mathrm{inverse}, \mathrm{solve},
\log\det\}$, all consumers draw from a single
\texttt{lu\_factor}---the factorization reuse JAX's
\texttt{solve}/\texttt{slogdet} adjoints get implicitly---which closed
the measured GP gap to JAX from $1.40\times$ to $1.24\times$ at $n{=}1024$.
Misses degrade performance, never correctness; retargeting is a second
registry.

\todopaper{Tech-mapping phase 2 (DP covering over the cell library,
using per-cell cost) and a GPU tier (Triton \cite{triton} emission for fused affine
regions; a second cell registry) are designed but not landed; the
evaluation marks where they matter.}

\section{Evaluation}\label{sec:eval}

All numbers: Apple M-series CPU, 2 threads, fp32, PyTorch 2.7; JAX
baselines are CPU \texttt{jit} with the same thread pinning, timed
min-of-$N$ with \texttt{block\_until\_ready}. Gradients
are verified against \texttt{torch.autograd} (typically
$\le 10^{-6}$ relative) and against an interpreter oracle; program
rewrites additionally against exact $\mathbb{F}_p$ fingerprints. The
project's 900-test suite includes a differential fuzzing harness
(480 random expressions per run, compiler vs.\ interpreter vs.\
ground truth, compared under name alignment) and machine-verified
gradients for every fused cell.

\subsection{Derived backward passes vs.\ runtime autograd}

\begin{table}
\caption{Loss+gradient microbenchmarks at $(256,1024)$, CPU (2 threads,
min-of-40). The tensorgrad column is a gradient \emph{derived} from a
three-line primitive definition and compiled; ``torch fused'' is the
framework's built-in hand-written kernel under autograd
(\texttt{layer\_norm}/\texttt{gelu}/\texttt{softmax}); ``torch prim.'' is
autograd through the \emph{same} naive primitive composition tensorgrad
starts from.}
\label{tab:micro}
\small
\begin{tabular}{lrrr}
\toprule
grad workload & tensorgrad & torch fused & torch prim. \\
\midrule
layer-norm & 0.24\,ms & 0.22\,ms & 0.48\,ms \\
gelu (tanh) & 1.27\,ms & 1.28\,ms & 1.06\,ms \\
softmax & 0.45\,ms & 0.47\,ms & 0.34\,ms \\
conv1d fwd+bwd & \multicolumn{3}{c}{$1.5\times$ over torch (MNIST step, Table~\ref{tab:suite})} \\
CE--softmax HVP & \multicolumn{3}{c}{parity with double backward} \\
\bottomrule
\end{tabular}
\end{table}

The result (Table~\ref{tab:micro}) is the thesis in miniature, and the
honest reading is a \emph{match}, not a rout: across all three elementwise kernels the
derived-and-compiled gradient tracks torch's built-in hand-written
kernel to within measurement noise (layer-norm $0.24$ vs.\ $0.22$\,ms,
GELU $1.27$ vs.\ $1.28$, softmax $0.45$ vs.\ $0.47$)---having been
produced from a definition with no backward pass written. Against
autograd through that same primitive definition, the derived layer-norm
gradient is $2.0\times$ faster ($0.24$ vs.\ $0.48$\,ms), because
factoring collapses the mean/variance backward that eager autograd
leaves separate; for softmax and GELU, torch's eager kernels already
fuse the primitive backward well and the three land at parity. The claim
is therefore not a speedup over autograd in general---it is that a kernel
\emph{nobody wrote} reaches vendor-fused performance, and that where the
derivative has real structure to exploit (convolutions, and the
second-order workloads below) the head start becomes decisive.

\subsection{The founding benchmark: Hessians}\label{sec:hessian}

$\partial^2 \mathrm{CE}(\mathrm{softmax}(x), y)/\partial x^2$ simplifies,
in 0.23\,s, to the two-term
$\mathrm{diag}(s) - ss^{\top}$ (batched), \emph{with the target $y$
eliminated}---because $y$ was declared on the simplex
(\S\ref{sec:constraints}) and $\sum_v y = 1$ is a rewrite. Hessian-vector
products compile to parity with \texttt{torch} double-backward. No
mainstream AD system exposes this closed form; symbolic systems that
could do not compile it.

Where the closed form does asymptotically less work, the wall clock
follows. At $n{=}2048$, $\partial^2(x^{\top}Ax)$ compiles to literally
$A + A^{\top}$ (the tests pin a two-statement ceiling, $x$ never read)
and runs in 2.1\,ms, against 17.7\,ms for \texttt{torch.func.hessian} \cite{functorch}
and 42.7\,ms for \texttt{jax.hessian} ($8\times$ and $20\times$; a direct
head-to-head---the suite harness of Table~\ref{tab:suite} times the same
row end-to-end at the same ratios): both
frameworks replay autodiff at runtime, and no amount of XLA fusion helps
when the algorithm itself is $n$ backward passes. The pattern holds on a
real second-order training loop and grows with dimension: assembling the
logistic-regression Hessian $X^{\top}\mathrm{diag}(\sigma(1-\sigma))X$
at $(n,d)=(2048,256)$---derived, never written---is ONE GEMM at 0.5\,ms,
against 1.2\,ms for \texttt{jax.hessian}'s $d$-pass transform loop and
4.7\,ms for \texttt{torch.func} ($2.5\times$ and $9\times$, widening
with $d$; at toy $d{=}32$ JAX's dispatch still wins, which is why the
claim is about real sizes). The logsumexp Hessian
(softmax plus a rank-one correction) shows the boundary honestly: our
0.72\,ms beats \texttt{torch.func}'s 1.47\,ms, but XLA materializes the
dense $n^2$ result faster still (0.47\,ms)---once the output is
unavoidably quadratic, the win reverts from asymptotic to constant.

\subsection{Expectations without sampling; input sets as theorems}
\label{sec:vae}

An \texttt{Expectation} node resolves Gaussian expectations by Stein's
lemma \cite{stein1981} during simplification, and composing it with derived derivatives
yields a model no runtime-AD framework can express
(\texttt{examples/vae.py}): a VAE \cite{vae} whose ELBO is written as its
definition, $\mathbb{E}_z[\log p(x|z)] - \mathbb{E}_z[\log q(z|x) -
\log p(z)]$ with $z \sim \mathcal{N}(\mu(x), \mathrm{diag}\,\sigma^2)$,
and resolves entirely to closed form. The textbook Gaussian KL,
$\tfrac12\sum(\mu^2 + \sigma^2 - 2\log\sigma - 1)$, appears nowhere in
the program text: it is \emph{derived} (verified against Monte-Carlo
estimates and against the textbook form). There is no reparametrization
trick and no sampling---$z$ never exists at runtime---so gradients have
zero variance. Training needs no optimizer either: the resolved ELBO is
quadratic in each weight block, so $\partial^2/\partial W^2$ \emph{of a
derived expectation} yields exact per-block Newton solves (the 4-edge
Hessian of a matrix block factors as a Kronecker product whose factors
are recovered from the derived Hessian itself by two Delta
contractions), and cycling the four blocks reaches the analytic
optimum---probabilistic PCA \cite{ppca}---in five rounds and 0.2\,s wall
(one-time compile included; the steady-state round is the 10\,ms quoted
below). No step size exists anywhere in the file. Head to head on time-to-optimum (reaching
$1.05\times$ the analytic PCA floor): the exact method takes 3 rounds and
10\,ms; the reparametrization-trick estimator torch and JAX must use
instead---runtime autodiff cannot differentiate through sampling, and both
hand-code the closed-form KL that tensorgrad derives---needs ${\sim}800$
Adam \cite{adam} steps and 160--190\,ms, a $16$--$19\times$ gap, with gradient noise
the exact method simply does not have.

One observation from this exercise generalizes. The compiled Newton
update for the posterior standard deviation consumes only
$(\sigma, W_{\mathrm{dec}})$: cancellation removed the data and the
encoder from the program, because a linear-Gaussian posterior's
covariance is input-independent---a theorem of PPCA that the derived
gradient rediscovered. (We first investigated it as a compiler bug: an
input had ``disappeared''. Every stage checked out.) Together with the
exactly-zero attention key-bias gradients of \S\ref{sec:adjoint}, a
pattern emerges that we did not design for: \emph{the input set of a
compiled program is a machine-derived statement of conditional
independence}, visible before a single number is computed. Runtime
autodiff cannot make such statements; it evaluates at a point. The same
discipline powers the rest of the example suite: a Gaussian process \cite{gp}
whose marginal-likelihood gradients derive the cookbook identities \cite{matrixcookbook}
$\partial \log|K| = \mathrm{tr}(K^{-1}\,\partial K)$ and
$\partial(K^{-1}) = -K^{-1}(\partial K)K^{-1}$ from the \texttt{det} and
\texttt{inverse} rewrite rules (matching finite differences to
$10^{-10}$), a Newton optimizer that is a single compiled expression
with quadratic convergence to $|g|^2 \approx 10^{-13}$, and a DDPM \cite{ddpm} whose
forward corruption process is algebra inside the training step.

\subsection{End-to-end: a GPT with no autograd}\label{sec:e2e}

\begin{table}
\caption{minGPT \cite{mingpt} (3 blocks, $d{=}48$) on the sorting task; 2-thread CPU,
min-of-$N$ on a quiet machine. The tensorgrad step program computes loss,
all 53 gradients, and the full AdamW update (160 outputs) as one compiled
function; \texttt{torch.set\_grad\_enabled(False)} is global. ``+
cells'' opts the same program into the fused attention/layer-norm/GELU/
AdamW cells of \S\ref{sec:techmap}.}
\label{tab:e2e}
\small
\begin{tabular}{lr}
\toprule
training step & time \\
\midrule
tensorgrad, all-derived (eager) & 42\,ms \\
tensorgrad, all-derived ($\times$ Inductor) & 25\,ms \\
tensorgrad + cells ($\times$ Inductor) & 7.4\,ms \\
torch autograd + SDPA + fused AdamW (compiled) & 5.2\,ms \\
JAX \texttt{jit} of the whole step & 4.5\,ms \\
\bottomrule
\end{tabular}
\end{table}

Both tensorgrad variants train to 100\% held-out accuracy (peak RSS
657\,MB eager vs.\ $\sim$285\,MB for torch). The gap structure is now
measured rather than conjectured. Between the all-derived and cell
variants lies the elementwise tail of the derived backward---the price
of full generality, recoverable kernel by kernel through the cell
registry. The remaining ${\approx}1.5\times$ to JAX is \emph{architectural}: we
emit into torch's backend, and XLA's whole-program fusion beats
torch-Inductor on this workload at every scale we measured (jax 4.8 vs.\
torch 5.2\,ms at $d{=}48$; the ratio holds at $d{=}384$, where both
are firmly compute-bound and tensorgrad+cells sits $1.2\times$ above
torch). A backend we rent bounds us by its own ceiling; the workloads of
\S\ref{sec:hessian}, where algebra changes the exponent rather than the
constant, are where the symbolic layer wins outright.

\subsection{Compile-time behavior}

Deriving, normalizing and lowering loss+53 gradients: 0.5\,s; planning
plus code generation on the first call: 5.5\,s (all one-time; earlier in
development this took 44\,s---the hash-consed fingerprints of
\S\ref{sec:iso} and the memoized structural recursions of
\S\ref{sec:depth} are what moved it). The adjoint pass runs in
0.5--6.8\,s at all scales of Table~\ref{tab:adjoint}. Equality
infrastructure: the exact-verification fallback fired zero times across
the loss+53-gradient build ($8{,}736$ queries; \S\ref{sec:iso}). One
systems footnote: a process that
compiles many programs (our benchmark suite compiles 15+) trips
TorchDynamo's per-frame recompile ceiling, which under
\texttt{fullgraph=True} raises rather than falling back; the runtime
raises the ceiling and keeps a per-program graceful fallback.
Everything in this paragraph is one-time cost; steady-state steps run
straight-line code.

\subsection{What is measured, and what is missing}\label{sec:eval-todo}

The three-framework comparison is a maintained benchmark suite
(\texttt{benchmarks/}): ten benchmark rows (nine models; GPT appears
both all-derived and $+$cells), each with a
tensorgrad, a torch, and (where expressible) a JAX side, every module
gated at import by a cross-framework correctness check on losses and
gradients. Table~\ref{tab:suite} is one run of the compiled mode.

\begin{table}
\caption{The benchmark suite, compiled mode, one consistent run
(tensorgrad under Inductor; torch under \texttt{torch.compile} as
written; JAX under \texttt{jit}), min-of-15 per row in ms;
\textbf{bold} = fastest of the three. The GPT rows use the suite
harness, whose torch step graph-breaks on its own \texttt{.item()}
logging; the dedicated quiet-machine comparison of
Table~\ref{tab:e2e} is authoritative there. conv's torch step does not
compile (dynamo error), so its baseline is elided.}
\label{tab:suite}
\small
\begin{tabular}{lrrr}
\toprule
step & tg & torch & jax \\
\midrule
conv (MNIST) & 2.2 & -- & $\mathbf{1.6}$ \\
diffusion (DDPM, AdamW) & 0.65 & 0.69 & $\mathbf{0.32}$ \\
embed (word2vec, scatter) & 2.2 & 2.3 & $\mathbf{1.6}$ \\
GP marginal likelihood & 0.20 & 0.39 & $\mathbf{0.13}$ \\
GPT (all-derived) & 19.1 & 9.5 & $\mathbf{6.2}$ \\
GPT (+cells) & 8.4 & 9.6 & $\mathbf{6.2}$ \\
Hessian $x^{\top}Ax$ & $\mathbf{1.6}$ & 17.6 & 43.1 \\
MLP (residual, AdamW) & 3.2 & 2.4 & $\mathbf{1.4}$ \\
Newton--GLM step & $\mathbf{0.59}$ & 21.8 & 1.2 \\
VAE (exact vs.\ reparam) & 0.46 & 0.30 & $\mathbf{0.15}$ \\
\bottomrule
\end{tabular}
\end{table}

The rows sort themselves by the paper's thesis. Where the algebra
changes the \emph{algorithm}, tensorgrad is fastest outright---the two
second-order rows: the Hessian row is the closed form of
\S\ref{sec:hessian}, and the Newton row solves with one derived GEMM
where \texttt{torch.func} replays $n{+}1$ backward passes and
\texttt{jax.hessian} a $d$-pass transform loop. Everywhere the program
is the \emph{same} algorithm, the per-step race is a constant-factor
contest that XLA's whole-program fusion wins on CPU: tensorgrad still
beats torch on several rows (GP $0.20$ vs.\ $0.39$ from the
\texttt{solve}/\texttt{slogdet}/LU cells; GPT$+$cells $8.4$ vs.\ $9.6$;
diffusion at parity) but trails JAX, and the honest gap is the backend,
not the algebra (\S\ref{sec:e2e}). The VAE row is the sharpest case of
why per-step time is the wrong axis: the exact method reaches the
optimum in 3 Newton rounds against ${\sim}800$ Adam steps
(\S\ref{sec:vae}), so its slower \emph{step} hides a $16$--$19\times$
faster \emph{convergence}---it is in the table anyway, because a suite
that showed only its flattering metric would defeat its purpose. What
remains unmeasured: XLA compile times next to ours, and any GPU
numbers.

\todopaper{\textbf{GPU tier} (future work): the evaluation is scoped to
ahead-of-time CPU compilation (stated in the abstract); Triton emission
for fused affine regions and A100/H100 numbers are a designed but
unlanded second cell registry.}

\todopaper{\textbf{Scale}: one training run beyond gpt-nano (12-block
GPT-2 dims now plans in 19.8\,GB and should execute on a 40\,GB device);
plus a second task domain (the conv/MNIST pipeline exists).}


\todopaper{\textbf{Formal statements}: a semantics appendix (expressions
as multigraphs, evaluation as a functor to multilinear maps; equality =
isomorphism soundness/completeness), termination of the rewrite system
under the declared measure, and the Schwartz--Zippel error bounds for
verification and consolidation (union bound over rewrites; the refusal
gate's independence argument).}

\section{Related Work}\label{sec:related}

\paragraph{Autodiff systems.} JAX \cite{jax}, PyTorch \cite{pytorch}, and
Enzyme \cite{enzyme} differentiate programs at runtime or at the IR of a
general language \cite{griewank2008}; Tapenade \cite{tapenade} and Zygote
\cite{zygote} likewise transform source or SSA; derivatives are computed, not simplified. Dex \cite{dex} and named-tensor
proposals \cite{namedtensor,einops} share our index discipline without the
algebraic layer. Our adjoint pass can be read as: reverse mode is a
\emph{theorem about contraction orderings} that survives being forgotten
and rediscovered by a compiler.

\paragraph{Computer algebra.} Mathematica \cite{mathematica} and \texttt{sympy} \cite{sympy} rewrite
expression trees with AC-matching and search;
\S\ref{sec:language} argues for quotient-by-representation instead.
Mathematica's \texttt{PossibleZeroQ} anticipates our use of randomized
identity testing \cite{schwartz1980}; we push it from a predicate to load-bearing compiler
verification.

\paragraph{Tensor compilers.} TACO \cite{taco}, TVM \cite{tvm}, Tensor Comprehensions
\cite{tensorcomprehensions}, and XLA/Inductor \cite{xla,pytorch2}
optimize loop nests or kernel graphs below the level of algebra, in the
scheduling-language lineage of Halide \cite{halide}; we sit above them and, on CPU, emit into one of them
(Inductor) as our kernel generator. Equality saturation \cite{eqsat} (egg \cite{egg}; Tensat \cite{tensat})
offers globally optimal rewriting at memory cost; our design keeps the
language layer deterministic and search-free and quarantines
cost-directed transformation in compiler passes---the adjoint pass shows
a case where the winning transformation is found structurally, with no
search at all.

\paragraph{EDA.} Layout assignment and cell selection are deliberate
transplants of placement and technology mapping \cite{dagon,abc}; functional (rather than
structural) matching via value fingerprints mirrors Boolean matching \cite{booleanmatching} in
logic synthesis.

\paragraph{Contraction ordering.} Planning inside our einsum nodes
follows the \texttt{opt\_einsum} lineage \cite{opteinsum,cotengra};
optimal ordering is NP-hard in general \cite{lam1997np}. Our adjoint
pass is complementary: it fixes the \emph{global} orientation of a
gradient graph, after which per-node planning is local.

\section{Conclusion}

A tensor language whose equality is isomorphism, whose derivatives are
rewrites, and whose compiler treats reverse-mode AD, numerical stability,
and kernel selection as recoverable \emph{properties of algebra} can
train a real model with autograd switched off---and, on the workloads
where its symbolic head start matters, outrun the tape.

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\appendix
\theoremstyle{remark}\newtheorem{remark}{Remark}[section]

%% The theorem-like environments (definition, lemma, theorem, proposition,
%% remark) and the proof environment used below are provided by acmart
%% (which loads amsthm). If this appendix is compiled outside acmart, add:
%%   \theoremstyle{definition}\newtheorem{definition}{Definition}[section]
%%   \theoremstyle{plain}\newtheorem{theorem}{Theorem}[section]
%%   \newtheorem{lemma}[theorem]{Lemma}\newtheorem{proposition}[theorem]{Proposition}
%%   \theoremstyle{remark}\newtheorem{remark}[theorem]{Remark}
\providecommand{\sem}[1]{\left[\!\left[#1\right]\!\right]}
\newcommand{\Free}{\operatorname{free}}
\newcommand{\Fp}{\mathbb{F}_{p}}
\newcommand{\dg}{\mathsf{d}}

\section{Formal Semantics}\label{app:semantics}

This appendix discharges the obligations stated in
\S\ref{sec:eval-todo}. Part~\ref{app:denot} gives the denotational
semantics---an expression as a labeled multigraph and the evaluation
functor $\sem{\cdot}$ into finite-dimensional multilinear maps---and
proves that isomorphic graphs denote equal maps (Soundness), then
delimits Completeness for the multilinear fragment.
Part~\ref{app:term} exhibits the well-founded measure that the
deterministic language-layer rewrites decrease and proves strong
normalization. Part~\ref{app:sz} states the Schwartz--Zippel identity-%
testing bound used by verification and consolidation, the union bound
over a compilation, and the independence argument that makes the
fresh-seed refusal gate's undetected-miscompile probability negligible.

Throughout, $\mathrm{Sym}$ is the set of symbolic size names (\texttt{batch},
\texttt{seq}, $\dots$), a \emph{dimension assignment} is a function
$\delta:\mathrm{Sym}\to\mathbb{N}_{>0}$, and $R$ is a fixed commutative
ring (the reals at runtime, $\Fp$ under verification). Every edge $e$
carries a size symbol $\dim(e)\in\mathrm{Sym}$; under $\delta$ it ranges
over the index set $I_e^{\delta}=\{0,\dots,\delta(\dim(e))-1\}$.

\subsection{Denotational semantics}\label{app:denot}

\subsubsection*{Expressions as labeled multigraphs}

\begin{definition}[Expression graph]\label{def:graph}
The \emph{graph} $G(e)$ of an expression $e$ is a finite labeled
multigraph with half-edges.
\begin{itemize}
\item \textbf{Vertices} are the primitive occurrences of the nine node
  types (\S\ref{sec:language}): \texttt{Variable}, \texttt{Delta},
  \texttt{Zero}, \texttt{Sum}, \texttt{Product}, \texttt{Function},
  \texttt{Derivative}, \texttt{Rename}, and scalar constants. Each vertex
  $v$ carries a \emph{label} $\ell(v)$ recording its type together with
  all data that participates in identity but is not itself a wire (the
  variable name, a \texttt{Delta}'s size symbol, \texttt{Sum} weights,
  a \texttt{Function} signature). This is exactly the $\textit{label}$
  component of the \textit{structure}() datum of \S\ref{sec:iso}.
\item \textbf{Edges} (internal) are the \emph{named indices} shared
  between two endpoints: a contracted edge of a \texttt{Product}, a
  hyperedge of a \texttt{Delta}, a broadcast wire of a \texttt{Sum}.
  Formally an edge is a \emph{junction}: an equivalence class of node--%
  port endpoints identified as one wire. Its interior name is used only
  to group endpoints and is then discarded, so $\alpha$-renaming of a
  contracted index is invisible in $G(e)$.
\item \textbf{Half-edges} (dangling) are the \emph{free} indices
  $\Free(e)$: endpoints of a junction with a distinguished
  \texttt{free} port and no matching interior endpoint. These are the
  expression's output indices; $\sem{e}$ is a tensor over exactly this
  index set.
\end{itemize}
An isomorphism $G(e_1)\cong G(e_2)$ is a bijection of vertices and of
junctions preserving vertex labels, incidence, and port roles;
\emph{edge-preserving} isomorphism additionally fixes the naming of the
half-edges in $\Free$.
\end{definition}

\begin{definition}[Equality]
$e_1\equiv e_2$ iff $G(e_1)\cong G(e_2)$ by an edge-preserving
isomorphism. This is the relation decided by the fingerprint-plus-VF2
procedure of \S\ref{sec:iso}; commutativity and associativity of
\texttt{Product}/\texttt{Sum} and $\alpha$-renaming of contracted edges
are not rewrites but consequences of $\cong$, because \texttt{Product}
and \texttt{Sum} hold unordered multisets and interior edge names are
quotiented in Definition~\ref{def:graph}.
\end{definition}

\subsubsection*{The evaluation functor}

Fix $\delta$ and a \emph{valuation} $\rho$ assigning to each
\texttt{Variable} $v$ with free edges $\Free(v)$ a tensor
$\rho(v)\in R^{\prod_{e\in\Free(v)} I_e^{\delta}}$ of matching shape.
Write $T^{\delta}_{S}=R^{\prod_{e\in S} I_e^{\delta}}$ for the space of
tensors with free-edge set $S$; an element of $T^{\delta}_S$ is
equivalently a multilinear map $\prod_{e\in S}(R^{I_e^\delta})^{*}\to R$,
which is the target category.

\begin{definition}[Evaluation]\label{def:eval}
$\sem{\cdot}_{\delta,\rho}$ sends an expression $e$ with
$\Free(e)=S$ to an element $\sem{e}_{\delta,\rho}\in T^{\delta}_{S}$,
defined by structural recursion. Writing an index tuple as $\vec\imath$
and $\vec\imath|_A$ for its restriction to edge set $A$:
\begin{align*}
\sem{v}_{\vec\imath} &= \rho(v)_{\vec\imath}, \\[-1pt]
\sem{\texttt{Zero}}_{\vec\imath} &= 0, \\[-1pt]
\sem{\Delta_n(e_1,\dots,e_k)}_{i_1,\dots,i_k}
   &= \begin{cases} 1 & i_1=\cdots=i_k,\\ 0 & \text{otherwise,}\end{cases}
   \qquad (k=0:\ \sem{\Delta_n}=\delta(n)\in R),\\[-1pt]
\sem{\textstyle\sum_t w_t\, e_t}_{\vec\imath}
   &= \textstyle\sum_t w_t\,\sem{e_t}_{\vec\imath}
     \qquad (\text{terms broadcast to a common }S),\\[-1pt]
\sem{\textstyle\prod_t e_t}_{\vec\imath}
   &= \sum_{\vec\jmath\in \prod_{c\in C} I_c^{\delta}}\;
      \prod_t \sem{e_t}_{(\vec\imath,\vec\jmath)\,|_{\Free(e_t)}},\\[-1pt]
\sem{\mathrm{Rename}_{\pi}(e)}_{\vec\imath}
   &= \sem{e}_{\vec\imath\circ\pi},\\[-1pt]
\sem{f(e_1,\dots,e_m)}
   &= \widehat{f}\big(\sem{e_1},\dots,\sem{e_m}\big),
\end{align*}
where $C$ is the set of internal (contracted) edges of the product,
$\pi$ is the free-edge bijection carried by the \texttt{Rename}, and
$\widehat f$ is the multilinear (or, in general, analytic) map named by
the \texttt{Function} signature. \texttt{Derivative} is not given a
direct clause: it denotes the derivative map of its operand and is
eliminated by the rewriting of \S\ref{sec:autodiff} before evaluation
(the eager interpreter refuses to evaluate an unresolved
\texttt{Derivative}; cf.\ \texttt{extras/evaluate.py}).
\end{definition}

The two structural clauses are the paper's slogans made precise:
\emph{contraction is summation over shared edges} (the \texttt{Product}
clause is exactly a named-index einsum), and \emph{$\Delta$ is the
identity/diagonal/copy tensor} (order $2$: the identity matrix; order
$1$: the all-ones vector; order $0$: the scalar $\delta(n)$, the value
returned by \texttt{evaluate} for an edgeless \texttt{Delta}).

\begin{lemma}[$\Delta$ is the copy tensor]\label{lem:delta}
For any edge set the family $\{\sem{\Delta_n(\cdots)}\}$ satisfies the
Frobenius/copy laws: contracting a leg of $\Delta_n(e_1,\dots,e_k)$ with a
leg of $\Delta_n(e_1',\dots,e_{k'}')$ yields $\Delta_n$ of arity
$k+k'-2$, and contracting an order-$2$ $\Delta_n$ against a tensor
renames the touched edge. Hence a \texttt{Delta} vertex acts as a
hyperedge fusing all its incident junctions into one wire, which is
precisely the single-junction \textit{structure}() declaration of
\S\ref{sec:iso}.
\end{lemma}

\begin{proof}
Immediate from $\sem{\Delta_n}_{\vec\imath}=[\,\text{all }i\text{ equal}\,]$:
a contraction $\sum_j [\![i_1{=}\cdots{=}j]\!]\,[\![j{=}\cdots{=}i_{k'}]\!]$
collapses to $[\![\text{all remaining equal}]\!]$, and
$\sum_j[\![i{=}j]\!]\,x_{j}=x_i$. Thus every leg indexes the same value,
so the incidence is a single equivalence class of endpoints, i.e.\ one
junction.
\end{proof}

\subsubsection*{Soundness}

\begin{theorem}[Soundness of equality]\label{thm:sound}
If $e_1\equiv e_2$ then $\sem{e_1}_{\delta,\rho}=\sem{e_2}_{\delta,\rho}$
for every dimension assignment $\delta$ and valuation $\rho$.
\end{theorem}

\begin{proof}
By induction on the structure of the isomorphism
$\Phi:G(e_1)\xrightarrow{\ \cong\ }G(e_2)$, using that
Definition~\ref{def:eval} is a fold over the same
\textit{structure}() data on which $\cong$ is defined
(\S\ref{sec:iso}: ``one source of structural truth''). $\Phi$ maps the
root of $e_1$ to a root of $e_2$ of equal label, hence of equal node
type and equal non-wire data, and restricts to isomorphisms of the
child subgraphs; the induction hypothesis equates their denotations.
It remains to check each clause is invariant under the residual freedom
in $\Phi$:
\begin{itemize}
\item \texttt{Product}: $\Phi$ permutes the (unordered) factor multiset
  and bijects the internal edges $C_1\to C_2$. The clause
  $\sum_{\vec\jmath}\prod_t(\cdots)$ is invariant under permuting the
  factors (commutativity of the product in $R$) and under relabeling the
  bound summation indices $\vec\jmath$ (a change of summation variable),
  which is exactly the freedom $\Phi$ has on $C$. Associativity is not a
  separate case: the flattened multiset carries no bracketing.
\item \texttt{Sum}: $\Phi$ permutes the term multiset while preserving
  the weight attached to each term (the weight is part of the child
  role/label), so $\sum_t w_t(\cdots)$ is preserved by commutativity of
  $+$.
\item \texttt{Delta}: by Lemma~\ref{lem:delta} the value is a symmetric
  function of its legs, so any permutation of them induced by $\Phi$
  fixes it.
\item \texttt{Rename}: an edge-preserving $\Phi$ carries $\pi$ to the
  matching free-edge relabeling, and $\sem{\mathrm{Rename}_\pi(e)}$ is
  $\sem e$ precomposed with that bijection.
\item \texttt{Variable}, \texttt{Function}, \texttt{Zero}: determined by
  the label alone (name / signature / type), which $\Phi$ preserves.
\end{itemize}
Since every clause is invariant, $\sem{e_1}=\sem{e_2}$.
\end{proof}

\begin{remark}
Soundness is what licenses the fingerprint as a decision procedure for
inequality (unequal denotations $\Rightarrow$ non-isomorphic
$\Rightarrow$ distinct fingerprints, contrapositive of
Theorem~\ref{thm:sound} composed with fingerprint completeness on
$\cong$), and is the property the differential fuzzing harness of
\S\ref{sec:sz} tests empirically.
\end{remark}

\subsubsection*{Completeness for the multilinear fragment}

Let the \emph{multilinear fragment} $\mathcal{M}$ be the sublanguage
built from \texttt{Variable}, \texttt{Delta}, \texttt{Rename},
\texttt{Sum}, and \texttt{Product} (no opaque \texttt{Function}, no
\texttt{Derivative}). Every $e\in\mathcal{M}$ denotes, for fixed
$\delta$, a map $\rho\mapsto\sem{e}_{\delta,\rho}$ that is a
\emph{multilinear polynomial}: multilinear in the entries of each
\texttt{Variable} block and of total degree at most the number of
variable-leaves along any contraction path.

Raw graph isomorphism is \emph{not} complete on $\mathcal{M}$: an
order-$2$ $\Delta_n$ inserted on an edge, a $\texttt{Zero}$ summand, or a
folded weight changes $G(e)$ without changing $\sem e$
(Lemma~\ref{lem:delta}). These are exactly the semantic---not
syntactic---equalities. The language therefore normalizes: let
$\mathrm{nf}(e)$ be the normal form produced by the deterministic
rewrites of Part~\ref{app:term} (identity-$\Delta$ absorption, $\Delta$
fusion, \texttt{Zero} elimination, weight folding, flattening). We state
completeness relative to normal forms.

\begin{proposition}[Completeness on normal forms]\label{prop:complete}
Assume $R$ is an infinite integral domain (or $|R|$ exceeds the degree
bound). For $e_1,e_2\in\mathcal{M}$ with $\Free(e_1)=\Free(e_2)$:
\[
  \big(\forall\rho\ \sem{e_1}_{\delta,\rho}=\sem{e_2}_{\delta,\rho}\big)
  \iff \mathrm{nf}(e_1)\equiv \mathrm{nf}(e_2).
\]
\end{proposition}

\begin{proof}[Proof sketch]
($\Leftarrow$) is Soundness (Theorem~\ref{thm:sound}) together with the
semantics-preservation of each normalization rule (each clause of
Part~\ref{app:term} is an instance of an $R$-algebra identity, checked
once per rule). ($\Rightarrow$) is coefficient extraction. A
$\Delta$-reduced $\mathcal M$-normal form is a \texttt{Sum} of
\texttt{Product} \emph{monomials}, each monomial a contraction pattern
that pairs specified variable-slots; its denotation is the multilinear
polynomial whose coefficient tensor records, for each choice of one
entry per participating variable, which index identifications the pattern
imposes (a product of $\Delta$'s). Over an infinite domain a multilinear
polynomial is determined by its coefficient tensor (a polynomial that
vanishes at all points of an infinite domain is the zero polynomial), and
the coefficient tensor determines the monomial's contraction pattern up
to the symmetries already quotiented by $\cong$; distinct monomials are
linearly independent, so equal polynomials have equal monomial multisets,
i.e.\ isomorphic normal forms. \qedhere
\end{proof}

\begin{remark}[Why the field must be large, and the bridge to Part C]
Proposition~\ref{prop:complete} needs an infinite (or large) domain
precisely at coefficient extraction. Over a small finite ring two
distinct multilinear polynomials can agree everywhere, so isomorphism of
normal forms is \emph{no longer} complete for pointwise equality. This
is exactly the regime of the compiler's value numbering: after the
adjoint and factoring passes (\S\ref{sec:adjoint}) many subgraphs denote
equal maps through non-isomorphic normal forms that the language layer
does not reach. Part~\ref{app:sz} recovers a probabilistic completeness
for these by testing the polynomial identity at random points of a large
finite field $\Fp$---the randomized backstop for the equalities that
structural $\cong$ cannot see.
\end{remark}

\subsection{Termination of the rewrite system}\label{app:term}

We show the deterministic language-layer rewrite relation
$\rightarrow_{\mathcal L}$ (the simplification catalog of
\S\ref{sec:rules-as-data} together with the constraint engine of
\S\ref{sec:constraints}) is \emph{strongly normalizing}: no expression
admits an infinite chain of rewrites. This substantiates the
``terminating rewriting toward a normal form'' claim of
\S\ref{sec:language}; confluence (hence uniqueness of normal forms) is
the separate, empirically tested obligation of the confluence-harness
\textsc{todo}, discussed in the closing remark.

\subsubsection*{The measure}

For an expression $e$ and a vertex $v\in G(e)$, write $\lvert
\mathrm{sub}(v)\rvert$ for the number of vertices \emph{strictly below}
$v$ (its operand subterm size). Define three non-negative integer
potentials:
\begin{align*}
\Phi_\partial(e) &= \!\!\sum_{v:\ \ell(v)=\texttt{Derivative}}\!\!
   \big(1+\lvert\mathrm{sub}(v)\rvert\big),
   &\text{(unresolved differentiation)}\\
\Phi_\rho(e) &= \!\!\sum_{v:\ \ell(v)=\texttt{Rename}}\!\!
   \big(1+\lvert\mathrm{sub}(v)\rvert\big),
   &\text{(pending lazy renames)}\\
N(e) &= \#\{\text{vertices of } G(e)\}.
   &\text{(node count)}
\end{align*}
The \emph{termination measure} is the lexicographic triple
\[
  \mu(e)=\big(\Phi_\partial(e),\ \Phi_\rho(e),\ N(e)\big)\in\mathbb{N}^3,
  \qquad \text{ordered by } <_{\mathrm{lex}}.
\]
$(\mathbb{N}^3,<_{\mathrm{lex}})$ is well-founded (a finite lexicographic
product of well-founded orders), so it suffices to show every rule
strictly decreases $\mu$.

\subsubsection*{Rule schemata and their effect on $\mu$}

The rules partition into three groups; we exhibit the strict decrease in
the responsible lexicographic coordinate and non-increase in the higher
ones.

\emph{(1) Derivative pushing} (\S\ref{sec:autodiff}): linearity over
\texttt{Sum}, the product rule over \texttt{Product}, the chain rule at
\texttt{Function}, and the base rules at leaves (a leaf $\Delta$ or a
signature derivative). A \texttt{Derivative} whose operand $u$ has
subterm size $s=\lvert\mathrm{sub}(v)\rvert$ is replaced by
\texttt{Derivative} nodes applied only to \emph{proper} subterms of $u$
(the summands of a \texttt{Sum}; the factors $t_i$ in
$\partial\prod_i t_i=\sum_i(\prod_{j\ne i}t_j)\,\partial t_i$; the
arguments at a \texttt{Function}), whose subterm sizes sum to at most
$s-1$. Hence the contribution $(1+s)$ of $v$ is replaced by a total
$\le s$, and every base rule deletes a \texttt{Derivative} outright, so
$\Phi_\partial$ strictly decreases. These rules introduce no
\texttt{Derivative} above where one was not already, and although the
product rule may \emph{increase} $N$, that coordinate is dominated in
$<_{\mathrm{lex}}$.

\emph{(2) Rename pushing} (\S\ref{sec:depth}, lazy \texttt{Rename}): a
\texttt{Rename} over a $k$-ary node becomes $\le k$ \texttt{Rename}s over
its children (each of strictly smaller subterm size), and a
\texttt{Rename} at a \texttt{Variable}/\texttt{Delta} leaf is absorbed
into the leaf's edge names. By the same accounting $\Phi_\rho$ strictly
decreases. These rules create no \texttt{Derivative} node, so
$\Phi_\partial$ is unchanged.

\emph{(3) Shrinking rules}: identity-$\Delta$ absorption and $\Delta$
fusion (Lemma~\ref{lem:delta}), \texttt{Zero} elimination, weight and
constant folding, \texttt{Sum}/\texttt{Product} flattening, and
duplicate collection. Each strictly reduces $N$ and creates neither a
\texttt{Derivative} nor a \texttt{Rename}, so $\Phi_\partial,\Phi_\rho$
are unchanged and $N$ strictly decreases.

\emph{Constraint rewrites} (\S\ref{sec:constraints}) are a distinguished
family of group~(3). A declared equation replaces a subnetwork
isomorphic to its left-hand side $L$ by its right-hand side $R$, subject
to the \emph{declaration-time obligation} $N(R)<N(L)$ (``the right-hand
side must be strictly smaller (node count) than the left''). Because a
constraint's $R$ is a value pattern containing no \texttt{Derivative} and
no \texttt{Rename}, the rewrite leaves $\Phi_\partial,\Phi_\rho$ fixed and
strictly decreases $N$ by $N(L)-N(R)>0$.

\begin{theorem}[Strong normalization]\label{thm:sn}
$\rightarrow_{\mathcal L}$ is strongly normalizing: for every rewrite
$e\rightarrow_{\mathcal L} e'$, $\mu(e')<_{\mathrm{lex}}\mu(e)$; hence no
infinite rewrite sequence exists and every expression has at least one
normal form.
\end{theorem}

\begin{proof}
By the case analysis above, each rule strictly decreases one coordinate
of $\mu$ and does not increase any higher coordinate: group~(1) decreases
$\Phi_\partial$; group~(2) fixes $\Phi_\partial$ and decreases
$\Phi_\rho$; group~(3), including constraints, fixes
$(\Phi_\partial,\Phi_\rho)$ and decreases $N$. Thus every step strictly
decreases $\mu$ in $<_{\mathrm{lex}}$. An infinite rewrite chain would
induce an infinite strictly-descending sequence in
$(\mathbb{N}^3,<_{\mathrm{lex}})$, contradicting well-foundedness.
\end{proof}

\begin{remark}[Termination versus confluence]
Theorem~\ref{thm:sn} gives \emph{existence} of normal forms. Their
\emph{uniqueness}---the normal-form claim of \S\ref{sec:language}---is
confluence, which for a strongly normalizing system reduces by Newman's
lemma to \emph{local} confluence: joinability of the two reducts of each
critical pair. That is precisely the property the confluence harness of
\S\ref{sec:rules-as-data} fuzzes (overlapping redexes, both application
orders, canonical-equality of the results, with the Schwartz--Zippel test
of Part~\ref{app:sz} separating a genuine confluence failure from a
soundness bug). Until that harness is landed the uniqueness claim is
empirical; termination is unconditional.
\end{remark}

\subsection{Schwartz--Zippel bounds for verification and consolidation}
\label{app:sz}

The compiler evaluates every IR node exactly over a finite field
$\Fp$ at seeded random points (\texttt{compiler/szfp.py}). We record the
error analysis for the two uses of this machinery: pinning individual
rewrites (\S\ref{sec:sz}, verification) and merging value-equal subgraphs
under the fresh-seed refusal gate (\S\ref{sec:sz}, consolidation).

\subsubsection*{The field and the atoms formulation}

The prime is the Mersenne prime
\[
   p \;=\; 2^{19}-1 \;=\; 524287
   \qquad(\texttt{P} \text{ in } \texttt{szfp.py}).
\]
It is chosen so pairwise einsum steps stay \emph{exact} in
\texttt{int64}: residues are $<2^{19}$, so products are $<2^{38}$, and up
to $2^{24}$ summed terms per output element keep partial sums
$<2^{62}<2^{63}$ (asserted at runtime). ``Evaluated over $\Fp$'' is
therefore literal, not floating---no rounding can corrupt the test.

A trial draws (i) a dimension assignment $\delta$ with
$\delta(n)\in\{2,\dots,5\}$ keyed by symbol \emph{name}, and (ii) for each
\texttt{Variable} an i.i.d.\ uniform tensor over $\Fp$, keyed by
$(\text{seed},\text{trial},\text{name},\delta)$ rather than by program
identity (so separately lowered programs share points). Non-polynomial
atoms---$\exp,\log,\tanh,\mathrm{erf}$, fractional \texttt{pow},
$\mathrm{softmax}$, \texttt{gather}, opaque cells---are \emph{not}
evaluated; each is replaced by a fresh uniform field element keyed by its
operator and the value-fingerprints of its inputs. Under this
\emph{atoms formulation}, the polynomial/rational skeleton (einsums,
linear combinations, integer powers, affine indicators, structural
constants) is the object of the identity test, with the atoms as
additional free indeterminates.

\subsubsection*{Per-check bound}

\begin{lemma}[Schwartz--Zippel \cite{schwartz1980,zippel1979,demillo1978}]
\label{lem:sz}
Let $Q\in\Fp[X_1,\dots,X_m]$ be a nonzero polynomial of total degree
$\dg$ and let $\vec X$ be drawn uniformly from $S^m$ with $S\subseteq\Fp$.
Then $\Pr[Q(\vec X)=0]\le \dg/|S|$. For a rational identity the same
bound applies to the numerator of the difference over the locus where
denominators are nonzero.
\end{lemma}

\begin{theorem}[Per-check false-accept probability]\label{thm:percheck}
Let $e,e'$ be two IR fragments with the same free-edge set that are
\emph{unequal} as skeleton rational functions of the atoms and variable
entries, of combined degree $\le\dg$. A single $k$-trial fingerprint
check declares them equal with probability at most
\[
   \Big(\tfrac{\dg}{p}\Big)^{k}.
\]
\end{theorem}

\begin{proof}
The difference $e-e'$ is a nonzero rational function of degree $\le\dg$.
Each trial evaluates it at a point whose variable- and atom-coordinates
are drawn uniformly and independently over $\Fp$ (points that hit a zero
denominator are resampled with a fresh salt, i.e.\ we condition on the
evaluation being defined; this only removes a $O(\dg/p)$ fraction of the
domain and the bound of Lemma~\ref{lem:sz} holds on the conditional to
first order). By Lemma~\ref{lem:sz} one trial fails to witness the
inequality with probability $\le\dg/p$. The $k$ trials use independent
seeds, so the events are independent and the joint probability is
$(\dg/p)^{k}$.
\end{proof}

The degrees arising in a lowered program are small---bounded by the
number of variable-leaves on a contraction path---so in practice
$\dg/p\approx 10^{-5}$ (the figure cited in \texttt{szfp.py}), and at the
default $k=3$ a single check has false-accept probability
$\approx(10^{-5})^{3}=10^{-15}$. The test is \emph{one-sided}: a true
skeleton identity \emph{always} passes (fingerprints of equal rational
functions are equal by construction), so the only error is a false
accept, and \emph{only} for skeleton (not analytic-atom) inequalities;
an equality that holds only through an analytic identity of the atoms
(e.g.\ $\exp(x)/\sum\exp(x)$ vs.\ $\mathrm{softmax}(x)$) is a documented
false-negative, never a false accept.

\subsubsection*{Union bound over a compilation}

\begin{theorem}[Verification of a compile]\label{thm:union}
Suppose a compilation applies rewrites $r_1,\dots,r_R$, each an equality
claim pinned by an independent $k$-trial check
(Theorem~\ref{thm:percheck}), with per-rewrite degree $\le\dg$. The
probability that \emph{any} unsound rewrite escapes detection is at most
\[
   R\,\Big(\tfrac{\dg}{p}\Big)^{k}.
\]
\end{theorem}

\begin{proof}
Let $B_i$ be the event that $r_i$ is unsound yet its check passes. By
Theorem~\ref{thm:percheck}, $\Pr[B_i]\le(\dg/p)^{k}$ (a sound rewrite
contributes $0$). The undetected-miscompile event is $\bigcup_i B_i$;
the union bound gives $\Pr[\bigcup_i B_i]\le\sum_i\Pr[B_i]\le
R(\dg/p)^{k}$.
\end{proof}

For a 3-block GPT step the compiler touches on the order of
$R\sim10^{3.5}$ compute nodes ($2{,}925\to844$; $7{,}127$ emitted
kernels, \S\ref{sec:sz}); with $\dg/p\approx10^{-5}$ and $k=3$ the whole-%
compile bound is $R(\dg/p)^{k}\lesssim 10^{4}\cdot10^{-15}=10^{-11}$.

\subsubsection*{The fresh-seed refusal gate}

Consolidation (\S\ref{sec:sz}) merges subgraphs whose evaluations agree
at the $k$ points of a seed $\sigma$ under one consistent axis
permutation, rewiring consumers through permuted views. The merge is a
\emph{proposal}: it can be wrong either through a skeleton collision at
$\sigma$ or through a bookkeeping bug (a mispermutation, or a dropped
broadcast wire of the kind that produced the longest-lived value bug in
the compiler's history, \S\ref{sec:sz}). Any such bug changes the
program's value as a rational function. The \emph{refusal gate} then
re-evaluates the original program $O$ and the consolidated program $C$
side by side at a \emph{fresh} seed $\sigma'$, independent of $\sigma$,
and rejects the entire rewrite---reverting to $O$---on any mismatch.

\begin{theorem}[Refusal-gate soundness]\label{thm:gate}
Suppose consolidation produces $C$ from $O$ and that $C$ is a miscompile:
$\sem{C}\ne\sem{O}$ as skeleton rational functions, of degree $\le\dg$.
Let the gate use $k'$ trials at a seed $\sigma'$ drawn independently of
the merge seed $\sigma$. Then the gate fails to refuse with probability
at most $(\dg/p)^{k'}$, \emph{regardless of how the merge was chosen}.
Moreover, if the miscompile arose from a skeleton collision that fooled
the proposal stage at $\sigma$, the probability that the merge is both
proposed and survives the gate is at most
\[
   \Big(\tfrac{\dg}{p}\Big)^{k}\cdot\Big(\tfrac{\dg}{p}\Big)^{k'}
   \;=\;\Big(\tfrac{\dg}{p}\Big)^{k+k'}.
\]
\end{theorem}

\begin{proof}
$O-C$ is a nonzero degree-$\le\dg$ rational function. The gate's $k'$
points are drawn i.i.d.\ over $\Fp$ at seed $\sigma'$; by
Theorem~\ref{thm:percheck} they all fail to witness $O\ne C$ with
probability $\le(\dg/p)^{k'}$. This bound is unconditional on $\sigma$:
the gate's randomness is disjoint from the proposal's, so ``$O,C$ agree
at $\sigma'$'' is independent of everything that happened at $\sigma$.
For the compounded bound, let $A$ = ``the wrong merge is proposed'' (it
requires the merged subgraphs to collide at $\sigma$, so
$\Pr[A]\le(\dg/p)^{k}$ by Lemma~\ref{lem:sz} applied at $\sigma$) and $E$
= ``the gate accepts'' with $\Pr[E]\le(\dg/p)^{k'}$. Because $\sigma$ and
$\sigma'$ are independent seeds, $A$ and $E$ are independent, so
$\Pr[A\cap E]=\Pr[A]\Pr[E]\le(\dg/p)^{k+k'}$.
\end{proof}

\begin{remark}[Why the seed must be fresh]
Independence is the whole content of the gate. If the gate reused the
merge seed $\sigma$, then a skeleton collision that fooled the proposal
would fool the gate at the \emph{same} points with certainty: the two
events would be perfectly correlated, $E\supseteq A$, and the gate would
add \emph{zero} protection. Re-drawing at an independent $\sigma'$
replaces that correlation with a product, turning the residual error from
$(\dg/p)^{k}$ into $(\dg/p)^{k+k'}$ for collision-induced bugs and
bounding \emph{all} value-changing bugs---including deterministic
bookkeeping bugs, for which $\Pr[A]=1$ but $\sem C\neq\sem O$ still
holds---by the unconditional $(\dg/p)^{k'}$. At $k'=3$ and
$\dg/p\approx10^{-5}$ this is $\approx10^{-15}$ per compile, so the gate
converts any unsound merge into a clean no-op fallback with overwhelming
probability rather than a miscompile. This is the argument that
\S\ref{sec:sz} advances for making the gate the default contract between
\emph{all} passes, not only consolidation: the bound of
Theorem~\ref{thm:gate} depends on nothing specific to value numbering,
only on the pass changing the program's value.
\end{remark}

\end{document}
